---
title: "Final Project"
author: "Eric Han, Minghao Sun, Qifan Yang, Steven Luo, Zihan Jin"
date: "2024-04-06"
output: 
  pdf_document: default
---

```{r setup, include=FALSE}
library(tidyverse)
library(knitr)

opts_chunk$set(warning = FALSE, message = FALSE)
```

# Data cleaning
```{r}
country_indicators <- 
  read_csv("country_indicators.csv") %>%
  select(-...1) %>%  # remove first column
  select(iso3, everything()) %>%  # reorder the columns to put iso3 as column 1
  rename(country_code_iso3 = iso3)  # rename first column to country_code_iso3
```

```{r}
country_code <- read.csv("country_codes.csv") %>% # read the file
  select(ISO.alpha3.Code..M49., everything()) %>%  # reorder the columns to put iso3 as column 1
  rename(country_code_iso3 = ISO.alpha3.Code..M49.) %>% # rename first column to country_code_iso3
  rename(continent = Region.Name_en..M49.) %>% # rename the column representing continent name
  rename(country_name = Country.or.Area_en..M49.) %>% # rename the column representing country names
  rename(sub_region_name = Sub.region.Name_en..M49.) %>%  # rename the column representing sub-region names
  select(country_code_iso3, country_name, continent, sub_region_name)
```

```{r}
project_data <- inner_join(x=country_code, y=country_indicators, by="country_code_iso3") %>%
  rename(under5_mortality_rate = "sowc_child-mortality__under-five-mortality-rate_2021") %>% 
  rename(delivery_care = "sowc_maternal-and-newborn-health__delivery-care-2016-2021-r_skilled-birth-attendant") %>%
  rename(mcv = "sowc_child-health__intervention-coverage_immunization-for-vaccine-preventable-diseases-2021-j_mcv1") %>%
  rename(hepb = "sowc_child-health__intervention-coverage_immunization-for-vaccine-preventable-diseases-2021-j_hepb3") %>%
  rename(hib = "sowc_child-health__intervention-coverage_immunization-for-vaccine-preventable-diseases-2021-j_hib3") %>%
  rename(adolescent_mortality = "sowc_adolescent-health__adolescent-mortality-rate-2021_aged-10-19_total") %>%
  
  
  select(country_code_iso3, country_name, continent, sub_region_name, hdr_le_2021, under5_mortality_rate, delivery_care, mcv, hepb, hib, adolescent_mortality)

head(project_data)
```

```{r}
data <- project_data %>%
  filter(!is.na(hdr_le_2021))
  
head(data)
```

```{r}
q2_indicators <- project_data %>%
  filter(!is.na(delivery_care)) %>%
  filter(continent == "Asia")

head(q2_indicators)
```

```{r}
q3_indicators <- project_data %>%
  filter(!is.na(mcv), !is.na(hepb), !is.na(hib), !is.na(adolescent_mortality)) %>%
  filter(continent == "Asia")

head(q3_indicators)
```

# Research Question 1

### Introduction: Understanding global disparities in life expectancy is essential for public health planning, resource allocation, and policy formulation. As one of the most direct indicators of health and well-being, life expectancy mirrors the general health status within a territory/country. By definition from world data bank, life expectancy at birth indicates the number of years a newborn infant would live if prevailing patterns of mortality at the time of its birth were to stay the same throughout its life. To interpret how Asia's overall health and well-being is, it is imperative to understand the life expectancy of newborns on this continent. Hence, we want to investigate the estimated range of life expectancy in Asia and compare the mean of life expectancy in Asia to the rest of world.

### Question: What is the estimated range of mean life expectancy in Asian countries? Is the mean life expectancy in Asian countries consistent with the rest of the world?

# Bootstrapping (to investigate the estimate of life expectancy)

(1) Sort the data related to expectancy into two different tables for bootstrapping
```{r}
le_data <- data %>%
  mutate(is_asian = case_when(
    continent == "Asia" ~ "YES",
    continent != "Asia" ~ "NO"
  ))

asian_life_expectancy_data <- le_data %>% 
  filter(is_asian == "YES")

non_asian_life_expectancy_data <- le_data %>%
  filter(is_asian == "NO")

asian_le_table <- tibble(asian_life_expectancy_data)
non_asian_le_table <- tibble(non_asian_life_expectancy_data)
```

(2) Create the random samples while making sure that all missing values are removed.
```{r}
set.seed(130)
asian_sample <- asian_le_table %>% sample_n(size = 40, replace = TRUE)
non_asian_sample <- non_asian_le_table %>% sample_n(size = 40, replace = TRUE)
```

(3) Initialize variables for the simulate
```{r}
num_repetitions <- 1000
asian_simulated_values <- rep(NA, num_repetitions)
non_asian_simulated_values <- rep(NA, num_repetitions)
```
 
(4) Run the simulation for asian countries
```{r}
n <- 40
repetitions <- num_repetitions
sim1 <- asian_simulated_values
sim2 <- non_asian_simulated_values
set.seed(130)

for (i in 1:repetitions)
{
  new_sim <- sample(asian_sample$hdr_le_2021 ,size = n, replace=TRUE)
  sim_mean <- mean(new_sim)
  sim1[i] <- sim_mean
}

for (i in 1:repetitions)
{
  new_sim <- sample(non_asian_sample$hdr_le_2021 ,size = n, replace=TRUE)
  sim_mean <- mean(new_sim)
  sim2[i] <- sim_mean
}

sim1 <- tibble(mean = sim1, is_asian = "Asia")
sim2 <- tibble(mean = sim2, is_asian = "Non-Asia")

simulation <- full_join(x = sim1, y = sim2)
```

(5) Plot the simulated means of life expectancy
```{r}

simulation %>% ggplot(aes(x = mean)) +
  geom_histogram(
    color = "black",
    fill = "grey",
    bins = 20
  ) +
  labs(x = "Average Life Expectancy in Asia and the rest of the world") +
  facet_wrap(~is_asian)

```
*From Observing and comparing the two histograms, we can find that the sampling distribution in Asia and the rest of the world are both symmetrical and unimodal with very similar spread of distribution.*

(6) Make a side-by-side boxplots to better observe the mean, IQR and outliers.
```{r}
simulation %>% ggplot(aes(x = is_asian, y = mean)) +
  geom_boxplot(
    color = "black",
    fill = "grey"
  ) +
  labs(x = "count", y = "average life expectancy") +
  coord_flip()
```

(7) Compute 95% confidence interval for Asian countries
```{r}
asia_lower_bound <- quantile(sim1$mean, p=0.025)
asia_upper_bound <- quantile(sim1$mean, p=0.975)

asia_lower_bound
asia_upper_bound
```
Base on the result above, we see that the 95% confidence interval for Asian countries are (70.70934, 73.87821)

And also, 95% confidence interval for non-Asian countries
```{r}
non_asian_lower_bound <- quantile(sim2$mean, p=0.025)
non_asian_upper_bound <- quantile(sim2$mean, p=0.975)

non_asian_lower_bound
non_asian_upper_bound
```
Based on the result above, we see that the 95% confidence interval for non-Asian countries are (67.33909, 72.51791)

It seems like that the average life expectancy in Asia is larger than the rest of the world from the result of boostrapping confidence intervals.
So next, we will conduct a two-sample hypothesis testing to find out if that is true.

# Hypothesis Testing 

(1) Declare the null hypothesis and alternative hypothesis
The null hypothesis of this hypothesis test is the average life expectancy in Asia is the same as the rest of the world. 
The alternative hypothesis of this hypothesis test is the average life expectancy in Asia is larger than the rest of the world. We set the alpha level $\alpha = 0.05$

$$H_0:\mu_a - \mu_w = 0 \ vs \ H_A: \mu_a - \mu_w > 0$$
where $\mu_a$ is the average life expectancy in Asia and $\mu_w$ is average life expectancy in other countries in the world.

(2) Compute the test statistic for the hypothesis test.
```{r}
test_stat <- le_data %>%
  group_by(is_asian) %>%
  summarise(means = mean(hdr_le_2021), .groups="drop") %>%
  summarise(value = diff(means)) %>%
  as.numeric()

print(test_stat)
```
(3) Run the simulation
```{r}
set.seed(130)
n_trials <- 1000
test_stat_simulations <- rep(NA, n_trials)

for(i in 1:n_trials){
  simdata <- le_data %>%
    mutate(is_asian = sample(is_asian), replace = FALSE)
  test_stat_sim <- simdata %>%
    group_by(is_asian) %>%
    summarise(means = mean(hdr_le_2021), .groups="drop") %>%
    summarise(value = diff(means)) %>%
    as.numeric()
  test_stat_simulations[i] <- test_stat_sim
}
```

(4) Plot the sampling distribution of difference in mean of life expectancy between Asia and the rest of the world.
```{r}
tibble(simulated_diff = test_stat_simulations) %>% 
  ggplot(aes(x = simulated_diff)) +
  geom_histogram(
    color = "black",
    fill = "grey"
  ) +
  labs(x = "Simulated Difference in Average Life Expectancy")
```
Finally, we calculate the p-value based on the null hypothesis $H_0$
```{r}
pvalue <- sum(test_stat_simulations >=  test_stat) / n_trials

print(pvalue)
```
The p-value of 0.039 is smaller than the alpha level 0.05, so there is moderate/strong evidence against the null hypothesis. As a result, we reject the null hypothesis in favour of the alternative that the average life expectancy in Asia is larger than the average life expectancy of the rest of the world.

# Conclusion

*In conclusion, the estimated range of life expectancy in Asia is (70.70934, 73.87821). It means that we are 95% confident that the true mean of life expectancy in Asia in 2021 fall between 70.70934 and 73.87821. To be more specific, by 95% confident, it means that 95% of the time, a 95% confidence interval would include the true mean of life expectancy in Asian countries. On the other hand, as a frame of context, the estimated range of life expectancy in non-Asian countries is (67.33909, 72.51791). It means that we are 95% confident that the true mean of life expectancy in non-Asian countries falls between 67.33909 and 72.51791. Through our hypothesis test, the data displayed moderate-to-strong evidence against our null hypothesis that the average life expectancy is the same as the rest of the world. It is safe to conclude that on average, Asian countries' life expectancy is larger than the rest of the world.*

*With this finding, we are able to proceed our research aimed to find what might have been associated with this larger life expectancy. Since Asia has a larger life expectancy as the rest of the world, by conducting further research on data of Asian countries, we are able to find valuable information contributing to this large life expectancy and use these information to help the rest of the world in working towards SDG Goal 3  .*


# Research Question 2

### Introduction: Reducing childhood mortality is a key universal health problem. We want to investigate whether we can lower the childhood mortality rate to improve the well-being in Asia by raising the delivery care rate.

### Question: Is there a linear association between the under-five mortality rate and the delivery care rate in Asia? If there is, what is the linear association?

### Definitions:(1) Delivery care rate: The proportion of births attended by skilled health personnel as defined as the percentage of live births attended by skilled health personnel (doctor, nurse, midwife).(2) Under-five mortality rate: Probability of dying between birth and exactly 5 years of age, expressed per 1,000 live births.

# Building model
(1) Visualize the association with scatter plots
```{r}
ggplot(data = q2_indicators, aes(y = under5_mortality_rate, 
                                   x = delivery_care, color = sub_region_name)) +
  geom_point(alpha = 0.8) + 
  labs(x = "Delivery care rate (2016 to 2021)", 
       y = "Under-five mortality rate (2021)")

ggplot(data = q2_indicators, aes(y = under5_mortality_rate, 
                                   x = delivery_care)) + 
  geom_point() + 
  labs(x = "Delivery care rate (2016 to 2021)",
       y = "Under-five mortality rate (2021)") + 
  facet_wrap(~sub_region_name)
```
*We can observe that there might be a negative linear relationship between two variables, and the relationships are different among different sub-regions of Asia.*

(2) Quantify the association by correlation of different sub-regions
```{r}
q2_indicators %>%
  group_by(sub_region_name) %>%
  summarise(correlation = cor(under5_mortality_rate, delivery_care))
```

*We notice that the correlations for different sub-regions vary and Central Asia's magnitude of r is close to zero, which indicates a weak strength of linear association. Hence, we decide to choose Central Asia as our baseline in the model.*

(3) Create a multiple linear regression model
(a) Build a model
$$Y_i = \beta_0 + \beta_1x_1,i + \beta_2x_2,i + \beta_3x_3,i + \beta_4x_4,i + \beta_5x_5,i + \epsilon_i$$
$Y_i$: the under-five mortality rate
$x_1,i$: delivery care rate(the proportion of births attended by skilled health personnel)
$x_2,i$: 1 when it refers to Eastern Asia, otherwise 0
$x_3,i$: 1 when it refers to South-eastern Asia, otherwise 0
$x_4,i$: 1 when it refers to Southern Asia, otherwise 0
$x_5,i$: 1 when it refers to Western Asia, otherwise 0
$\beta_0$: the under-five mortality rate relates to x = 0 (intercept)
$\beta_1$: the slope of the baseline
$\beta_2$: the coefficients of $x_2,i$
$\beta_3$: the coefficients of $x_3,i$
$\beta_4$: the coefficients of $x_4,i$
$\beta_5$: the coefficients of $x_5,i$

(b) multiple coefficient test(t-test)
We need to use hypothesis test to check every coefficients:(i = 1,2,3,4,5)
$$H_0: \beta_i = 0$$
$$H_0: \beta_i \neq 0$$
We set $\alpha$ level at 0.05

(c) Check p-value
```{r}
model_1 <- lm(under5_mortality_rate~delivery_care + sub_region_name, 
              data =  q2_indicators)
summary(model_1)
```
*Since the p-value for *$\beta_1$ *are significantly smaller than the alpha value we set, we have strong evidence against the null hypothesis that the slope is zero. Hence, we can conclude that there exists a linear relationship between the under-five mortality rate and the delivery care rate in Central Asian countries. Moreover, we notice that Southern Asia fails the t-test, which means that we fail to reject the Null hypothesis(*$\beta_4 = 0$*), which indicates that the slopes of the regression lines are not parallel for the particular variable Southern Asia, and other variables fit the parallel linear regression lines model well.*
*The multiple R-squared value of 0.75 suggests that, in Asia, 75% of the variance in the under-five mortality rate can be explained by the delivery care rate in the model. The remaining 25% of the variance is unexplained by the model and can be attributed to either variables not included in the model or inherent variability in the data.*

*We have equations for fitted linear regression lines:*

For Central Asia countries:
$$Y_i = 113.28 - 0.91 * delivery\,care\,rate$$
For Eastern Asia:
$$Y_i = 113.28 - 0.91 * delivery\,care\,rate - 13.94$$
For South-eastern Asia:
$$Y_i = 113.28 - 0.91 * delivery\,care\,rate - 11.49$$
For Southern Asia:
$$Fail\,to\,fit\,in\,parallel\,lines\,model$$
For Western Asia:
$$Y_i = 113.28 - 0.91 * delivery\,care\,rate - 10.61$$

(d) Visualize linear regression lines
```{r}
library(broom)
augment(model_1)
```
Since Southern Asia case fails the t-test, we remove the Southern Asia Case
```{r}
asia_indicators_modified <- q2_indicators %>% 
  filter(sub_region_name != "Southern Asia")
model_1_modified <- lm(under5_mortality_rate~delivery_care + sub_region_name, 
              data =  asia_indicators_modified)
library(broom)
augment(model_1_modified)
```

```{r}
ggplot(data = asia_indicators_modified, aes(y = under5_mortality_rate, 
                                            x = delivery_care, 
                                            color = sub_region_name)) + 
  geom_point(alpha = 0.5) + 
  geom_line(data=augment(model_1_modified),
            aes(y=.fitted, colour=sub_region_name), lwd = 1.5)
  labs(x = "Delivery care rate (2016 to 2021)",
       y = "Under-five mortality rate (2021)") +
  theme_minimal()
```

# Conclusion

*In conclusion, there is a negative linear relationship between the under-five mortality rate and the delivery care rate among Asian countries in 2021, while the association varies in different sub-regions.*

*For Central Asian, Southern Asian, South-eastern Asian, and Western Asian countries, increasing the delivery care rate can significantly decrease the under-five mortality rate.*

*However, in East Asia, the observation of a vertical distribution of Eastern Asia countries' data suggests that while there is a universally high rate of delivery care, there exists significant variability in the under-five mortality rate.This implies that there might be other factors that affect the under-five mortality rate in East Asia. To lower the under-five mortality rate and improve the well-being of children, these countries should shift their focus towards these factors, rather than delivery care rate. According to the research by Madhav Kumar Bhusa and Shankar Prasad Khanal, education of mother, size of child at birth, age of mother at childbirth, place of residence, and birth interval were the significant and most frequently observed factors of under-five mortality. Nevertheless, specific investigations focusing on the context of East Asian countries remain scarce, presenting an opportunity for future research.*


# Research Question 3
*Questions:Is there a relationship between adolescent mortality of age 10 - 19 and various vaccine (MCV, HepB, Hib) coverage in Asia? Using information based on these two variables in data, what is the linear regression between adolescent mortality and vaccine coverage? *

# Building model
(1) Linear regression between mcv and adolescent_mortality
(a) Quantify the association by correlation of different sub-regions
```{r}
q3_indicators %>% group_by(continent) %>%
  summarise(cor(x = mcv, y = adolescent_mortality)) # calculate the correlationship between mcv and adolescent_mortality
```

(b) Build a model
$${Y}_i = {\beta}_0 + {\beta}_1{x}_i + {\epsilon}_i$$

${Y}_i$: the adolescent mortality

${x}_i$: mcv coverage rate

${\beta}_0$: the adolescent mortality rate relates to x = 0

${\beta}_1$: slope parameter

${\epsilon}_i$: random error term for observation

(c) State the null and alternative hypotheses
$${H}_0:{\beta}_1 = 0 \:\: vs \:\:{H}_A:{\beta}_1 \neq 0$$
We set $\alpha$ level at 0.05

(d) Check p-value
```{r}
mcv_model <- lm(adolescent_mortality ~ mcv, data = q3_indicators) # build a simple linear regression
summary(mcv_model)$coefficients # find the coefficients
```
*Since the p-value of *$\beta_1$ = 5.764076e-03*is significantly smaller than the alpha value we set, we have strong evidence against the null hypothesis that the slope is zero. Therefore, we can conclude that there is a linear relationship between adolescent mortality and and mcv coverage rate in Asian countries.*

(e) Visualize linear regression lines
```{r}
q3_indicators %>% ggplot(aes(x = mcv, y = adolescent_mortality, color = continent)) + 
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "lm", se = FALSE) # plot the graph
```
Equation: $\hat{y} = 12.8 - 0.09x_i$

(2) Linear regression between hepb and adolescent_mortality
(a) Quantify the association by correlation of different sub-regions
```{r}
q3_indicators %>% group_by(continent) %>%
  summarise(cor(x = hepb, y = adolescent_mortality)) # calculate the correlationship between hepb and adolescent_mortality
```
(b) Build a model
$${Y}_i = {\beta}_0 + {\beta}_1{x}_i + {\epsilon}_i$$
${Y}_i$: the adolescent mortality

${x}_i$: hepb coverage rate

${\beta}_0$: the adolescent mortality rate relates to x = 0

${\beta}_1$: slope parameter

${\epsilon}_i$: random error term for observation

(c) State the null and alternative hypotheses
$${H}_0:{\beta}_1 = 0 \:\: vs \:\:{H}_A:{\beta}_1 \neq 0$$
We set $\alpha$ level at 0.05

(d) Check p-value
```{r}
hepb_model <- lm(adolescent_mortality ~ hepb, data = q3_indicators) # build a simple linear regression

summary(hepb_model)$coefficients # find the coefficients
```
*Since the p-value of *$\beta_1$ = 0.0346655585*is significantly smaller than the alpha value we set, we have strong evidence against the null hypothesis that the slope is zero. Therefore, we can conclude that there is a linear relationship between adolescent mortality and and hepb coverage rate in Asian countries.*

(e) Visualize linear regression lines
```{r}
q3_indicators %>% ggplot(aes(x = hepb, y = adolescent_mortality, color = continent)) + 
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "lm", se = FALSE) # plot the graph
```
$Equation: \hat{y} = 10.9 - 0.07x_i$

(2) Linear regression between hepb and adolescent_mortality
(a) Quantify the association by correlation of different sub-regions
```{r}
q3_indicators %>% group_by(continent) %>%
  summarise(cor(x = hib, y = adolescent_mortality)) # calculate the correlationship between hib and adolescent_mortality
```

(b) Build a model
$${Y}_i = {\beta}_0 + {\beta}_1{x}_i + {\epsilon}_i$$
${Y}_i$: the adolescent mortality

${x}_i$: hib coverage rate

${\beta}_0$: the adolescent mortality rate relates to x = 0

${\beta}_1$: slope parameter

${\epsilon}_i$: random error term for observation

(c) State the null and alternative hypotheses
$${H}_0:{\beta}_1 = 0 \:\: vs \:\:{H}_A:{\beta}_1 \neq 0$$
We set $\alpha$ level at 0.05

(d) Check p-value
```{r}
hib_model <- lm(adolescent_mortality ~ hib, data = q3_indicators) # build a simple linear regression

summary(hib_model)$coefficients # find the coefficients
```

(e) Visualize linear regression lines
```{r}
q3_indicators %>% ggplot(aes(x = hib, y = adolescent_mortality, color = continent)) + 
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "lm", se = FALSE) # plot the graph
```
$Equation: \hat{y} = 11.4 - 0.07x_i$
```{r}
library(broom)
augment(model_1)
```
```{r}
mcv_model <- lm(adolescent_mortality ~ mcv + hib + hepb + mcv * hib + mcv * hepb + hepb * hib + mcv * hepb * hib, data = q3_indicators) # build a simple linear regression
summary(mcv_model)
```

```{r}
q3_indicators %>% ggplot(aes(x = mcv + hib + hepb + mcv * hib + mcv * hepb + hepb * hib + mcv * hepb * hib , y = adolescent_mortality, color = continent)) + 
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "lm", se = FALSE) # plot the graph
```
Equation: 
$$\hat{y} = (3.555) + (-0.109)mcv_i + (31.19)hib_i + (-31.08)hepb_i + (-0.327)mcv_i * hib_i + (0.327)mcv_i*hepb_i + (0.004)hib_i * hepb_i + (-0.00004)mcv_i*hib_i*hepb_i$$

# Conclusion
*In conclusion, the inverse association between mortality and vaccination rates for MCV (measles vaccine), HepB (hepatitis B vaccine) and Hib (Haemophilus influenzae type b vaccine) among adolescents aged 10-19 years in Asia is a significant finding. It also reflects Asian countries' progress toward SDG Goal 3 about Good Health and Well-Being.*

*This correlation suggests that higher vaccine coverage is associated with lower adolescent mortality, suggesting the potential effectiveness of vaccination programs in reducing adolescent mortality in Asia. These findings underscore the importance of prioritizing and scaling up vaccination efforts to ensure broad coverage and access to these essential vaccines.*

*Furthermore, this correlation highlights the critical role of immunization in preventing vaccine-preventable diseases and their associated complications, which can help reduce mortality in adolescents. It emphasized the need for continued investment in vaccination campaigns, health infrastructure and education initiatives to promote vaccine acceptance and uptake among Asian communities.*

*Overall, these findings highlight the importance of maintaining and strengthening vaccination efforts as a cornerstone of public health strategies aimed at improving health outcomes and reducing mortality among adolescents in Asia.*


References:
(1) Databank. (n.d.). https://databank.worldbank.org/metadataglossary/world-development-indicators/series/SP.DYN.LE00.IN#:~:text=Life%20expectancy%20at%20birth%20indicates,the%20same%20throughout%20its%20life.
(2) Bhusal MK, Khanal SP. A Systematic Review of Factors Associated with Under-Five Child Mortality. Biomed Res Int. 2022 Dec 5;2022:1181409. doi: 10.1155/2022/1181409. PMID: 36518629; PMCID: PMC9744612.